---
title: "Visual Analysis on data of US Department of Labor's Office of Foreign Labor Certification, Employment and Training Administration"
author: "Yash Rajaram Bagayatkar"
date: "10/7/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this R file, I have aimed to create a comprehensive report with visualizations providing insights on the data provided by the US Department of Labor's Office of Foreign Labor Certification, Employment and Training Administration about the employer's request on the H1B non-immigrant workers. The information about the data's attributes is provided in the Appendix section at the bottom of this report

# Initialization

In this section, I have loaded the necessary libraries and the H1B data. I have also replaced any spaces in the column names with '.'

```{r initialize}
library(tidyverse)
library(data.table)
library(ggplot2)
library(dplyr)
#Reading the NYC 311 Data set
h1b<-fread("h1bdata.csv")
names(h1b)<-names(h1b) %>%
  stringr::str_replace_all("\\s",".")

```

# Tidying the Dataset

After initialization, I have performed cleaning of the data to deal with missing values and also made changes with respect to some columns depending on the need for analysis. Instead of adding mean, or default values for numeric or categorical missing values, I have removed those rows altogether as the missing rows are very less as compared to the length of the whole dataset. I have removed empty string values from EMPLOYEE_STATE, PW_UNIT_OF_PAY and WAGE_UNIT_OF_PAY. I have also converted all the PREVAILING_WAGE, WAGE_RATE_OF_PAY_FROM, WAGE_RATE_OF_PAY_TO values to Hourly Wages out of the possible categories of Yearly, Monthly, Weekly, Bi-Weekly and Hourly. I have done this so that there is consistency in the wage related analysis that I have performed in the later section.

```{r tidy}
sum(is.na(h1b))
nrow(h1b)
h1b<-na.omit(h1b)
sapply(h1b, class)
head(h1b)
h1b<-dplyr::filter(h1b,h1b$EMPLOYER_STATE!="")
h1b<-dplyr::filter(h1b,h1b$PW_UNIT_OF_PAY!="")
h1b<-dplyr::filter(h1b,h1b$WAGE_UNIT_OF_PAY!="")
h1b<-within(h1b,{
  temp<-PW_UNIT_OF_PAY=="Year"
  PREVAILING_WAGE[temp]<-PREVAILING_WAGE/(52*40)
  PW_UNIT_OF_PAY<-"Hour"
})
h1b<-within(h1b,{
  temp<-PW_UNIT_OF_PAY=="Month"
  PREVAILING_WAGE[temp]<-PREVAILING_WAGE*(12/(52*40))
  PW_UNIT_OF_PAY<-"Hour"
})
h1b<-within(h1b,{
  temp<-PW_UNIT_OF_PAY=="Week"
  PREVAILING_WAGE[temp]<-PREVAILING_WAGE/40
  PW_UNIT_OF_PAY<-"Hour"
})
h1b<-within(h1b,{
  temp<-PW_UNIT_OF_PAY=="Bi-Weekly"
  PREVAILING_WAGE[temp]<-PREVAILING_WAGE/80
  PW_UNIT_OF_PAY<-"Hour"
})

h1b<-within(h1b,{
  temp<-WAGE_UNIT_OF_PAY=="Year"
  WAGE_RATE_OF_PAY_FROM[temp]<-WAGE_RATE_OF_PAY_FROM/(52*40)
  WAGE_RATE_OF_PAY_TO[temp]<-WAGE_RATE_OF_PAY_TO/(52*40)
  WAGE_UNIT_OF_PAY<-"Hour"
})
h1b<-within(h1b,{
  temp<-WAGE_UNIT_OF_PAY=="Month"
  WAGE_RATE_OF_PAY_FROM[temp]<-WAGE_RATE_OF_PAY_FROM*(12/(52*40))
  WAGE_RATE_OF_PAY_TO[temp]<-WAGE_RATE_OF_PAY_TO*(12/(52*40))
  WAGE_UNIT_OF_PAY<-"Hour"
})
h1b<-within(h1b,{
  temp<-WAGE_UNIT_OF_PAY=="Week"
  WAGE_RATE_OF_PAY_FROM[temp]<-WAGE_RATE_OF_PAY_FROM/40
  WAGE_RATE_OF_PAY_TO[temp]<-WAGE_RATE_OF_PAY_TO/40
  WAGE_UNIT_OF_PAY<-"Hour"
})
h1b<-within(h1b,{
  temp<-WAGE_UNIT_OF_PAY=="Bi-Weekly"
  WAGE_RATE_OF_PAY_FROM[temp]<-WAGE_RATE_OF_PAY_FROM/80
  WAGE_RATE_OF_PAY_TO[temp]<-WAGE_RATE_OF_PAY_TO/80
  WAGE_UNIT_OF_PAY<-"Hour"
})
head(h1b)
```

# Exploring the Dataset

The exploring section deals with the core of this report highlighting observations made using the visualizations. I have performed series of visualizations for seasonal analysis, OOccupational analysis and geographical analysis in succession.

```{r explore1}
# Frequency per year
h1b %>%
  group_by(CASE_SUBMITTED_YEAR) %>%
  summarise(count=n())
h1b<-dplyr::filter(h1b,CASE_SUBMITTED_YEAR!=2011 & CASE_SUBMITTED_YEAR!=2012)
#Number of Yearly petition
h1b %>%
  group_by(CASE_SUBMITTED_YEAR) %>%
  summarise(count=n()) %>%
  ggplot(aes(x=CASE_SUBMITTED_YEAR,y=count,group=1)) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("Number of Petitions") +
  ggtitle("Number of Yearly petition")
```

I have removed the years 2011 and 2012 years from CASE_SUBMITTED_YEAR column as they amount to just a handful observations which do not contribute to a data with 530000 observations. In the above visualization, the drastic increase can be seen at the year 2016 and then a further increase 2017.

```{r explore2}
#H1b petitions every year as per Case_status
ggplot(h1b) +
  geom_bar(aes(x=CASE_SUBMITTED_YEAR,fill=CASE_STATUS)) +
  xlab("Year") +
  ylab("Cases Submitted per Year") +
  ggtitle("H1b petitions every year as per Case status")
```

The above visualization is performed to understand how the case statuses out of CERTIFIED, CERTIFIEDWITHDRAWN, DENIED AND WITHDRAWN are distributed along the years. We can see that primarily, the status for the cases in the dataset is certified.

```{r explore4}
# Seasonal analysis of Certified case status Petitions
Months<-c("January","February","March","April","May","June","July","August","September","October","November","December")
h1b$Case_Submitted_Month_Name<-Months[h1b$CASE_SUBMITTED_MONTH]
h1b.Certified<-dplyr::filter(h1b,h1b$CASE_STATUS=="CERTIFIED")
h1b.Certified%>%
  group_by(Case_Submitted_Month_Name)%>%
  summarise(count=n())%>%
  ggplot(aes(x=reorder(Case_Submitted_Month_Name,count),y=count,fill=reorder(Case_Submitted_Month_Name,-count)))+
  scale_fill_discrete(name="Months") +
  geom_bar(stat="identity") +
  xlab("CASE_SUBMITTED_MONTH") +
  ylab("Number of Petitions") +
  ggtitle("Monthly analysis of Certified Cases") +
  coord_flip()
```

For the next visualization, I have created a list of Month names as the CASE_SUBMITTED_MONTH column is numerical. The visualization shows a seasonal analysis throughout the years for every month. We can see from the above chart that H1B cases are primarily submitted by employers in the month of March followed by a steep decline on the month of February and then January. This states that the beginning of the year is when the maximum cases are submitted for H1B.

```{r explore3}
# Petitions per Occupation
h1b %>%
  group_by(SOC_NAME)%>%
  summarise(count=n())%>%
  arrange(desc(count))%>%
  head(10)%>%
  ggplot(aes(x=reorder(SOC_NAME,count),y=count,fill=reorder(SOC_NAME,-count)))+
  scale_fill_discrete(name="Occupations") +
  geom_bar(stat="identity") +
  xlab("Occupations") +
  ylab("Number of Petitions") +
  ggtitle("Petitions per Occupation") +
  coord_flip()
```

I have then moved towards an occupational analysis of the dataset. In the above dataset, I have shown the number of H1B petitions per occupation highlighting the top 10 occupations. The highest number in this case is of the field of computer related occupations. This overwhelming majority is followed by analysts and then engineers which are the in demand jobs for which employers apply for H1B status.

```{r explore5}
library(scales)
# Range Prevailing Wage for different Occupation
ggplot(h1b) +
  geom_boxplot(aes(PREVAILING_WAGE,SOC_NAME))+
  xlab("Prevailing Wages") +
  ylab("Occupations") +
  ggtitle("Prevailing Wages for different Occupation") +
  theme(axis.text=element_text(size=5.5),
        axis.title=element_text(size=14,face="bold")) +
  scale_x_continuous(labels=comma)
```

After seeing the number of petitions for each occupation, we can now move on to the wages where in the above visualization, we can see the various jobs and their boxplots showing the range for hourly prevailing wages. Although the previous chart leaders of Computer occupation, Analysts show variation, Engineers shown minimal variation in prevailing wages. On the contrary, considerable variation is shown for occupations in Marketing, and that of Scientists.

```{r explore8}
library(scales)
# Range of Wage Rate per Pay From (Minimum) for different Occupation
ggplot(h1b) +
  geom_boxplot(aes(WAGE_RATE_OF_PAY_FROM,SOC_NAME))+
  xlab("Wage Rate of Pay From") +
  ylab("Occupations") +
  ggtitle("Wage Rate of Pay From (Minimum Wage) for various Occupation") +
  theme(axis.text=element_text(size=5.5),
        axis.title=element_text(size=14,face="bold")) +
  scale_x_continuous(labels=comma)
```

Speaking of boxplots and wages, the next chart is also an occupational boxplot showing variation among the Beginning spectrum of the wage rate. Again, Computer occupation, Analysts show variation along with Scientists and Marketing occupations.

```{r explore9}
library(scales)
# Range of Wage Rate per Pay To (Maximum) for different Occupation
ggplot(h1b) +
  geom_boxplot(aes(WAGE_RATE_OF_PAY_TO,SOC_NAME))+
  xlab("Wage Rate of Pay To") +
  ylab("Occupations") +
  ggtitle("Wage Rate of Pay To (Maximum Wage) for different Occupation") +
  theme(axis.text=element_text(size=5.5),
        axis.title=element_text(size=14,face="bold")) +
  scale_x_continuous(labels=comma)
```

The last graph in the boxplot section is for the Wage Rate of Pay To column which is the higher side of spectrum for wage rates. The Computer occupation and Analysts do not show variation here hinting that there is a relatively constant wage rate for the higher end of the wage rate range. The scientists and marketing occupation still shows some variation in wage, but not drastic.

```{r explore6}
# USA Map for number of Petitions
library(usmap)
df1<-h1b%>%
  group_by(EMPLOYER_STATE)%>%
  summarise(count=n())
colnames(df1)[colnames(df1)=="EMPLOYER_STATE"]<-"state"
plot_usmap(regions=c("states"), data=df1,values="count",labels=TRUE,label_color="white")
```

Above is the map that I have used to show the geographical distribution of the number of H1B petitions where largely many states are seemingly almost evenly distributed distributed towards the lower side of the spectrum. However two light blue states show drastic increase in the number of petitions as compared to the other states and they are California and Texas followed by New Jeysey in the North East part of the United States of America.

```{r explore7}
#Average Prevailing wages per State

h1b%>%
  group_by(EMPLOYER_STATE)%>%
  summarise(Mean=mean(PREVAILING_WAGE))%>%
  head(10)%>%
  ggplot(aes(x=reorder(EMPLOYER_STATE,Mean),y=Mean,fill=reorder(EMPLOYER_STATE,-Mean)))+
  scale_fill_discrete(name="STATE") +
  geom_bar(stat="identity") +
  geom_text(aes(label=paste("$",round(Mean,1))),hjust=1.1) +
  xlab("STATE") +
  ylab("Average Wage") +
  ggtitle("Average Prevailing wages per State") +
  coord_flip()
```

Next, I have checked for the average prevailing wages per hour for the States with top 10 prevailing wages. It can be found that Arizona is leading this chart with an average prevailing wage of $69.6 .
The next highest prevailing wage is for the state of california with average of $65.7 . After this, Conneticut and Alabama average a preaviling wage of $47.5 and $41.1 per hour

# Conclusion

Using this report, I have attempted to solve some of intriguing questions with the help of powerful and insightful visualizations. It can be seen that US provides very good opportunity to many good employers to expand their work force by bringing in employees from all over the world using the H1B visa status. Even though states like California and Arizona have higher prevailing wage rates, states like Texas, New York and New Jersey also provide great opportunity for workers of different occupation, as shown by the number of petitions on the US Map visualization. I was also able to gain insights from the boxplots of the various occupation with varying range of prevailing wage, Wage rate to pay From and Wage rate to pay To. For the Scientists and Marketing employees even though varied greatly, along with more variance, there are more number of petitions for Computer related occupation, Analysts and Engineers. I have also performed seasonal analysis for the Certified status cases and found that most cases are sibmitted by the employer in the beginning of the year, with an overwhelming increase in the month of March followed by February and January.

# Appendix A: Data Dictionary for H1B Dataset

The dataset consists of the following columns:

1. CASE_STATUS: status associated with the last significant event of decision. Valid values include "Certified", "Certified-Withdrawn","Denied" and "Withdrawn"

2. CASE_SUMBITTED_DAY: The day the application was submitted

3. CASE_SUMBITTED_MONTH: The month the application was submitted

4. CASE_SUMBITTED_YEAR: The year the application was submitted

5. DECISION_DAY: The day on which the last significant event of decision was recorded

6. DECISION_MONTH: The month on which the last significant event of decision was recorded

7. DECISION_YEAR: The year on which the last significant event of decision was recorded

8. VISA_CLASS: Indicates the type of temporary application submitted for processing

9. EMPLOYER_NAME: Name of the employer submitting labor condition application

10. EMPLOYER_STATE: State of the employer submitting labor condition application

11. EMPLOYER_COUNTRY: Country of the employer submitting labor condition application

12. SOC_NAME: Occupation name associated with the job being requested for temporary labor condition

13. NAICS_CODE: Industry code associated with the employer requesting permanent labor condition

14. TOTAL_WORKERS: Total number of foreign workers requested by the employer

15. FULL_TIME_POSITION: Y=Full time position, N= Part time position

16. PREVAILING_WAGE: Prevailing wage for the job being requested for temporary labor

17. PW_UNIT_OF_PAY: Unit of pay. Valid values are "Year", "Month", "Week", "Bi-Weekly" and "Hour"

18. PW_SOURCE: Variables include: "OES", "CBA", "DBA", "SCA" or "Other"

19. PW_SOURCE_YEAR: Year the prevailing wage source was issued

20. PW_SOURCE_OTHER: If "Other" wage source, provide source of wage

21. WAGE_RATE_OF_PAY_FROM: Employer's proposed wage rate

22. WAGE_RATE_OF_PAY_TO: Maximum wage rate

23. WAGE_UNIT_OF_PAY: Unit of pay. Valid values are "Year", "Month", "Week", "Bi-Weekly" and "Hour"

24. H-1b_DEPENDENT: Y= Employer is H-1B dependent, N= Employer is not H-1B dependent

25. WILLFUL_VIOLATOR: Y= Employer has been previously found to be willful violator, N= Employer has not been a willful violator

26. WORKSITE_STATE: State information of the foreign worker's intended area of employment

27. WORKSITE_POSTAL_CODE: Zip code information of the foreign worker's intended area of employment